Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations

### 2015-08-10 docs/figures_tables/figure2.tex
Remade figures and added final comments before submission
Remade figures and sent revised manuscript to coauthors for comments

% Document
% Document
\documentclass[12pt, a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{authblk}
\usepackage{lipsum}
\usepackage{tikz}
\usepackage{hyperref}

% Figure and table formating
\usepackage{epsfig}
\usepackage{tabu}
\usepackage{rotating}
\usepackage{pbox}
\usepackage{framed, multicol}
\usepackage[framemethod=TikZ]{mdframed}

% Set up Frame
\mdfdefinestyle{MyFrame}{%
    linecolor=black,
    skipabove=10pt,
    outerlinewidth=2pt,
    roundcorner=20pt,
    innertopmargin=10pt,
    innerbottommargin=10pt,%\baselineskip,
    innerrightmargin=20pt,
    innerleftmargin=20pt,
    backgroundcolor=gray!50!white}

\usepackage{float}
\usepackage[left=1 in, right=1 in, top=1.25 in, bottom=1.25 in]{geometry}

% Fonts - Mathtime
%\usepackage{txfonts}
\usepackage{amsmath} % Add amssymb if not using Mathtime

% Text
\setlength{\parindent}{0.5in}
\frenchspacing  \tolerance = 800  \hyphenpenalty = 800

\usepackage{lineno} % Line numbers
\def\linenumberfont{\normalfont\footnotesize\ttfamily}
\setlength\linenumbersep{0.1 in}

\usepackage{setspace}

% Other
\usepackage{graphicx}
\usepackage[singlelinecheck=false,font=small,labelfont=bf]{caption}

@@

\begin{figure}

@@

    \node at (-3.2, 4.9) {A.};
    \node at (-6.7, 2.9) {B.};
    \node at (1.15, 2.9) {C.};

@@

\end{figure}

\centering

\textbf{Figure 1}

>>>>

\documentclass[tikz]{standalone}

\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{matrix}
\usetikzlibrary{backgrounds}

@@

@@

    \node at (-2.5, 3.8) {A.};
    \node at (-5, 2.2) {B.};
    \node at (1, 2.2) {C.};

@@

### 2015-08-10 docs/parasite_mortality_manuscript.tex
Remade figures and sent revised manuscript to coauthors for comments

Parasites can significantly impact animal populations by changing host
behavior, reproduction and survival. Detecting and quantifying these impacts is
critical for understanding disease dynamics and managing wild animal
populations. However, for wild hosts infected with macroparasites, it is
notoriously difficult to quantify the fatal parasite load and number of animals
that have died due to disease. When ethical or logistical constraints prohibit
experimental determination of these values, examination of parasite intensity
and distribution data may offer an alternative solution. In this study we
introduce a novel method for using intensity data to detect and quantify
parasite-induced mortality in wildlife populations. We use simulations to show
that this method is more reliable than previously proposed methods while
providing quantitative estimates of parasite-induced mortality from empirical
data that are consistent with previously published qualitative estimates.
However, this method, and all techniques that estimate parasite-induced
mortality from intensity data alone, have several important assumptions that
must be scrutinized before applying them to real-world data. Given that these
assumptions are met, our method is new exploratory tool that can help inform
more rigorous studies of parasite-induced host mortality.

@@

It is important to note that the flexibility of the negative binomial distribution may also reduce our ability to detect PIHM. If
a negative binomial can be fit to the observed post-mortality parasite
distribution then, regardless of how lethal the parasite was, it will be
impossible to detect PIHM because there is no need for a more complex model.
Many observed parasite distributions are well-fit by the negative binomial distribution \citep{Shaw1998}, suggesting
that systems where these methods are applicable without any a prior knowledge may be uncommon. However, if one has a prior knowledge about some aspect of the pre-mortality distribution \citep[e.g. assumes/knows the value of $k_p$,][]{Ferguson2011}, then the Likelihood Method could be applicable even if the the post-mortality distribution was well-fit by a negative binomial.

If one has evidence that the pre-mortality is not negative binomial, the generality of our method easily allows another distribution to be specified for $g(x, \boldsymbol{\phi})$. In particular, one could use the
resulting stationary host-parasite distribution from a stochastic host-parasite
model without parasite-induced host mortality \citep{Anderson1982a} to specify
the form of $g(x, \boldsymbol{\phi})$ and then apply the techniques discussed
in this paper to detect PIHM.  The general criteria necessary
for the Likelihood Method to detect PIHM in a stochastic host-parasite process
is that the stationary distribution of the process with
mortality is significantly different than the stationary distribution without mortality.
It is widely recognized that parasite-induced host mortality decreases the
aggregation of host-parasite distributions relative to those without mortality
\citep{Barbour2000}, suggesting that Likelihood Method could be applicable for
detecting PIHM in host-parasite systems experiencing stochastic parasite births, deaths, and immigrations in addition to host mortality.

If the Likelihood Method is applicable and the truncation of the negative binomial distribution is detected, one must be aware that the pattern may be caused by other
processes such as within host density dependence, age dependent variation in host
resistance and/or heterogeneous infection rates \citep{Anderson1982a,Rousset1996, McCallum2000a}. This means that in the event
that PIHM is detected, it may actually not be the result of PIHM.  Moreover, if host mortality depends on parasite intensity and additional variables (e.g. host sex, host size), failure to identify these important confounding variables could significantly affect the ability of these methods to correctly identify PIHM. However, both of these issues -- inferring process from pattern and confounding variables -- are well-recognized limitations of most statistical inference and are addressed via judicious model specification and selection \citep{Seber2003}.

@@

    \caption{A) Five potential shapes for a host-survival functions. In the simulations we used a gradual survival function (dotted line), and moderate survival function (dashed line), and a steep survival function (solid line). The linear and immediate survival functions represent two potential extremes that we do not include in the simulations. For each of these survival functions and the parameter combinations described in the main text, we tested the Type I error and power of the Likelihood (Like.) Method and Adjei Method. B) Gives the Type I error of each method over a range of pre-mortality sample sizes with a pre-mortality mean parasite intensity ($\mu_p$) of 50 and pre-mortality parasite aggregation ($k_p$) at 0.5. The red line shows the pre-set significance level of 0.05. C) Gives the power of each method for detecting PIHM over a range of post-mortality sample sizes for $\mu_p = 50$ and $k_p = 0.5$.  In general, the Likelihood Method has higher power and lower Type I error than the Adjei Method.  See the \emph{SI} 3 Fig 1 - 3 for Type I Error and power results for all parameter combinations.}

@@

    \caption{The power of the Likelihood Method (Like.) to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated. The curves were generated from 500 simulations for 10 pre-mortality sample sizes, $N_p$. The vertical, dotted-dashed lines indicate the sample size at which the power the for the Likelihood Method with steep and moderate survival functions is 0.8 (75 hosts for steep functions and 402 for moderate functions). The Likelihood Method with a gradual survival function never has a power above 0.8.}

>>>>

Parasites can significantly impact animal populations by changing host
behavior, reproduction and survival. Detecting and quantifying these impacts is
critical for understanding disease dynamics and managing wild animal
populations. However, for wild hosts infected with macroparasites, it is
notoriously difficult to quantify the fatal parasite load and number of animals
that have died due to disease. When ethical or logistical constraints prohibit
experimental determination of these values, examination of parasite intensity
and distribution data may offer an alternative solution. In this study we
introduce a novel method for using intensity data to detect and quantify
parasite-induced mortality in wildlife populations. We use simulations to show
that this method is more reliable than previously proposed methods while
providing quantitative estimates of parasite-induced mortality from empirical
data that are consistent with previously published qualitative estimates.
However, this method, and all techniques that estimate parasite-induced
mortality from intensity data alone, have several important assumptions that
must be scrutinized before applying them to real-world data. Given that these
assumptions are met, our method is a new exploratory tool that can help inform
more rigorous studies of parasite-induced host mortality.

@@

It is important to note that the flexibility of the negative binomial distribution may also reduce our ability to detect PIHM. If
a negative binomial can be fit to the observed post-mortality parasite
distribution then, regardless of how lethal the parasite was, it will be
impossible to detect PIHM because there is no need for a more complex model.
Many observed parasite distributions are well-fit by the negative binomial distribution \citep{Shaw1998}, suggesting
that systems where these methods are applicable without any \emph{a priori} knowledge may be uncommon. However, if one has \emph{a priori} knowledge about some aspect of the pre-mortality distribution \citep[e.g. assumes/knows the value of $k_p$,][]{Ferguson2011}, then the Likelihood Method could be applicable even if the the post-mortality distribution was well-fit by a negative binomial.

If one has evidence that the pre-mortality is not negative binomial, the generality of our method easily allows another distribution to be specified for $g(x, \boldsymbol{\phi})$. For example, one could use the
resulting stationary host-parasite distribution from a stochastic host-parasite
model without parasite-induced host mortality \citep{Anderson1982a} to specify
the form of $g(x, \boldsymbol{\phi})$ and then apply the techniques discussed
in this paper to detect PIHM.  The general criteria necessary
for the Likelihood Method to detect PIHM in a stochastic host-parasite process
is that the stationary distribution of the process with
mortality is significantly different than the stationary distribution without mortality.
It is widely recognized that parasite-induced host mortality decreases the
aggregation of host-parasite distributions relative to those without mortality
\citep{Barbour2000}, suggesting that the Likelihood Method could be generally applicable to host-parasite systems that follow the assumptions of many stochastic host-parasite models.  This is an intriguing area for further research.

If the Likelihood Method is applicable and the truncation of the negative binomial distribution is detected, one must be aware that the truncation pattern may be caused by other
processes such as within host density dependence, age dependent variation in host
resistance and/or heterogeneous infection rates \citep{Anderson1982a,Rousset1996, McCallum2000a}. This means that in the event
that PIHM is detected, it may actually not be the result of PIHM.  Moreover, if host mortality depends on parasite intensity and additional variables (e.g. host sex, host size), failure to identify these important confounding variables could significantly affect the ability of these methods to correctly identify PIHM. However, both of these issues -- inferring process from pattern and confounding variables -- are well-recognized limitations of most statistical inference and are addressed via judicious model specification and selection \citep{Seber2003}.

@@

    \caption{A) Five potential shapes for a host-survival functions. In the simulations we used a gradual survival function (dotted line), and moderate survival function (dashed line), and a steep survival function (solid line). The linear and immediate survival functions represent two potential extremes that we do not include in the simulations. For each of these survival functions and the parameter combinations described in the main text, we tested the Type I error and power of the Likelihood (Like.) Method and Adjei Method. B) Gives the Type I error of each method over a range of pre-mortality sample sizes with a pre-mortality mean parasite intensity ($\mu_p$) of 50 and pre-mortality parasite aggregation ($k_p$) at 0.5. The red line shows the pre-set significance level of 0.05. C) Gives the power of each method for detecting PIHM over a range of post-mortality sample sizes for $\mu_p = 50$ and $k_p = 0.5$.  In general, the Likelihood Method has higher power and lower Type I error than the Adjei Method.  See the \emph{SI} 3 Fig 1 - 3 for Type I error and power results for all parameter combinations.}

@@

    \caption{The power of the Likelihood Method (Like.) to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated. The curves were generated from 500 simulations for 10 pre-mortality sample sizes, $N_p$. The vertical, dotted-dashed lines indicate the sample size at which the power for the Likelihood Method with steep and moderate survival functions is 0.8 (75 hosts for steep functions and 408 for moderate functions). The Likelihood Method with a gradual survival function never has a power above 0.8.}
